---
title: "Running QC checks on MarineGEO data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running QC checks on MarineGEO data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(marinegeo.utils)
library(dplyr)
```

This vignette demonstrates the `marinegeo.utils` QC functions using
`seagrass_cover_example`, a multi-year seagrass percent cover dataset from
Indian River Lagoon, Florida (`partner_code = "USA-IRL"`).

There are two ways to run QC checks:

1. **`qc_run()`** — single entry point that runs all applicable tests for a
   known `table_id`, dispatching based on metadata.
2. **Individual check functions** — `qc_check_columns()`,
   `qc_check_data_types()`, and `qc_check_categorical_values()` — for targeted
   checks against a custom schema.

---

## Running all checks with `qc_run()`

`qc_run()` accepts a data frame (or file path) and a `table_id`. It queries
the internal `marinegeo_metadata` object to determine which columns are
expected, what data types they should have, and which columns have controlled
vocabularies — then runs all applicable tests.

```{r load-data}
data(seagrass_cover_example)
```

```{r qc-run}
result <- qc_run(
  seagrass_cover_example,
  table_id = "seagrass-cover-monitoring-v1"
)
```

The return value is a named list with four top-level elements:

```{r result-structure}
names(result)
```

`status` is the worst status across all tests (`"fail"` > `"warn"` > `"pass"`).
`n_rows` is the number of rows validated. `tests` is a named list with one
entry per test that was run.

```{r top-level}
cat("Overall status:", result$status, "\n")
cat("Rows validated:", result$n_rows, "\n")
cat("Tests run:", paste(names(result$tests), collapse = ", "), "\n")
```

---

## Inspecting individual test results

Each element of `result$tests` has the same structure: `test`, `status`,
`message`, `summary`, and `failures`.

### Column check

```{r column-check}
col_result <- result$tests$qc_check_columns

cat("Status:", col_result$status, "\n")
cat("Message:", col_result$message, "\n")
```

```{r column-failures}
col_result$failures
```

The example dataset is missing the `taxonomic_id` column. This column is
optional in practice (it is populated after a taxonomy join), but the schema
defines it as expected, so the check flags it as a failure.

### Data type check

```{r type-check}
type_result <- result$tests$qc_check_data_types

cat("Status:", type_result$status, "\n")
cat("Message:", type_result$message, "\n")
knitr::kable(type_result$summary)
```

All 16 present columns have the expected data types.

### Categorical values check

```{r cat-check}
cat_result <- result$tests$qc_check_categorical_values

cat("Status:", cat_result$status, "\n")
cat("Message:", cat_result$message, "\n")
knitr::kable(cat_result$summary)
```

The `summary` shows how many invalid values were found per column. Use
`failures` to see the specific rows:

```{r cat-failures-cover-code}
# Unique invalid values in cover_code
cat_result$failures |>
  filter(column_name == "cover_code") |>
  distinct(value)
```

`cover_code` has a value of `5`, which is not in the Braun-Blanquet scale
controlled vocabulary (`0`, `0.1`, `0.5`, `1`, `2`, `3`, `4`). `NA` values
are ignored by the categorical check.

---

## Extracting `[row, col]` coordinates from categorical failures

The `failures` data frame returned by `qc_check_categorical_values()` includes
both `row_index` and `col_index` — the exact `[row, col]` position of each
violating cell in the original data frame. Both are 1-based (R convention).

```{r row-col-coords}
df_bad <- data.frame(
  site_name = c("Site A", "Site B", "INVALID", "Site A"),
  cover_method = c("point count", "UNKNOWN", "point count", "Brawny-Blanquet"),
  stringsAsFactors = FALSE
)

rules_demo <- data.frame(
  column_name = c("site_name", "site_name", "cover_method", "cover_method"),
  value = c("Site A", "Site B", "point count", "Braun-Blanquet"),
  stringsAsFactors = FALSE
)

result_demo <- qc_check_categorical_values(df_bad, rules_demo)

knitr::kable(result_demo$failures)
```

`col_index` is the column's position in the original data frame
(`which(colnames(data) == column_name)`), so it remains correct when `data`
contains columns that are not covered by any rules.

To format these as compact `[row, col]` coordinate strings:

```{r row-col-format}
result_demo$failures |>
  mutate(coordinate = paste0("[", row_index, ", ", col_index, "]")) |>
  select(coordinate, column_name, value)
```

## Running individual checks

The individual check functions can be called directly with a custom schema,
without needing a registered `table_id`.

### `qc_check_columns()`

```{r check-columns-standalone}
result_cols <- qc_check_columns(
  data = seagrass_cover_example,
  expected_columns = c(
    "partner_code",
    "site_name",
    "sample_collection_date",
    "scientific_name",
    "percent_cover"
  )
)

cat("Status:", result_cols$status, "\n")
cat("Message:", result_cols$message, "\n")
```

### `qc_check_data_types()`

```{r check-types-standalone}
result_types <- qc_check_data_types(
  data = seagrass_cover_example,
  type_map = c(
    partner_code = "STRING",
    sample_collection_date = "DATE",
    transect = "TINYINT",
    percent_cover = "DOUBLE"
  )
)

cat("Status:", result_types$status, "\n")
cat("Message:", result_types$message, "\n")
```

### `qc_check_categorical_values()`

```{r check-cat-standalone}
rules <- data.frame(
  column_name = c("cover_method", "cover_method", "cover_method"),
  value = c("Braun-Blanquet", "point count", "percent bins")
)

result_cat <- qc_check_categorical_values(
  data = seagrass_cover_example,
  rules = rules
)

cat("Status:", result_cat$status, "\n")
cat("Message:", result_cat$message, "\n")
knitr::kable(result_cat$summary)
```

---

## Using `detail = FALSE` in automated pipelines

In automated pipelines where row-level detail is unnecessary (and memory use
matters), pass `detail = FALSE`. All tests still run and statuses are still
reported, but `failures` is `NULL` for every test.

```{r detail-false}
result_pipeline <- qc_run(
  seagrass_cover_example,
  table_id = "seagrass-cover-monitoring-v1",
  detail = FALSE
)

cat("Overall status:", result_pipeline$status, "\n")

# failures are suppressed
lapply(result_pipeline$tests, function(t) {
  cat(t$test, "-", t$status, "| failures:", is.null(t$failures), "\n")
})
```

---

## Reading from a file

`qc_run()` can read directly from a CSV or XLSX file path, avoiding the need
to pre-load data.

```{r file-path, eval = FALSE}
# CSV
result_csv <- qc_run(
  "path/to/seagrass_cover.csv",
  table_id = "seagrass-cover-monitoring-v1"
)

# Excel (specify sheet if needed)
result_xlsx <- qc_run(
  "path/to/seagrass_cover.xlsx",
  table_id = "seagrass-cover-monitoring-v1",
  sheet = 1L
)
```
