---
title: "Running QC checks on MarineGEO data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running QC checks on MarineGEO data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(marinegeo.utils)
library(dplyr)
```

This vignette demonstrates the `marinegeo.utils` QC functions using
`seagrass_cover_example`, a multi-year seagrass percent cover dataset from
Indian River Lagoon, Florida (`partner_code = "USA-IRL"`).

Every QC function returns the **same shape**: a `qc_issues` table with one row
per detected problem (zero rows means the data passed).
The columns are:

| column | meaning |
|---|---|
| `check` | the producing check, e.g. `"qc_check_numeric_ranges"` |
| `severity` | `"fail"` or `"warn"` |
| `row` | 1-based row number, or `NA` for table/column-level issues |
| `column` | offending column name, or `NA` |
| `col_index` | 1-based column position, or `NA` |
| `value` | the offending value, or `NA` |
| `issue` | machine-readable code, e.g. `"out_of_range"` |
| `message` | human-readable description |

Because it is just a tibble, you inspect and subset results with ordinary
`dplyr`. There are two ways to produce one:

1. **`qc_run()`** — single entry point that runs all applicable checks for a
   known `table_id`, dispatching based on metadata, and binds their issues into
   one table.
2. **Individual check functions** — `qc_check_columns()`,
   `qc_check_data_types()`, `qc_check_categorical_values()`,
   `qc_check_missing_values()`, `qc_check_numeric_ranges()`, and
   `qc_check_lookup_values()` — for targeted checks against a custom schema.

---

## Running all checks with `qc_run()`

`qc_run()` accepts a data frame (or file path) and a `table_id`. It queries
the internal `marinegeo_metadata` object to determine which columns are
expected, what data types they should have, and which columns have controlled
vocabularies — then runs all applicable checks.

```{r load-data}
data(seagrass_cover_example)
```

```{r qc-run}
result <- qc_run(
  seagrass_cover_example,
  table_id = "seagrass-cover-monitoring-v1"
)
```

The result prints as an ordinary tibble — one row per detected problem:

```{r print-result}
result
```

Run-level metadata is attached as attributes:

```{r top-level}
cat("Overall status:", attr(result, "status"), "\n")
cat("Rows validated:", attr(result, "n_rows"), "\n")
cat("Checks run:", paste(attr(result, "checks_run"), collapse = ", "), "\n")
```

`status` is the worst severity present (`"fail"` > `"warn"` > `"pass"`),
`n_rows` is the number of rows validated, and `checks_run` lists the checks
that executed for this `table_id`.

---

## Working with the issues table

Since the result is a tibble, summarising and filtering are plain `dplyr`.
A count by check and severity gives a quick overview:

```{r tally}
result |>
  count(check, severity) |>
  knitr::kable()
```

Filter to the rows for a single check to inspect them:

```{r filter-one-check}
result |>
  filter(check == "qc_check_categorical_values")
```

Or pull every fail-level issue across all checks:

```{r filter-fails}
result |>
  filter(severity == "fail") |>
  select(check, row, column, value, message)
```

---

## Inspecting individual checks within a run

Each check contributes rows tagged with its name in `check`.

### Column check

```{r column-check}
result |>
  filter(check == "qc_check_columns") |>
  select(severity, column, issue, message)
```

The example dataset is missing the `taxonomic_id` column. This column is
optional in practice (it is populated after a taxonomy join), but the schema
defines it as expected, so the check flags it as a `"fail"` with
`issue = "missing_column"`. Column-level issues have `row = NA`.

### Data type check

```{r type-check}
result |>
  filter(check == "qc_check_data_types")
```

A zero-row result means all present columns have the expected data types.

### Categorical values check

```{r cat-check}
result |>
  filter(check == "qc_check_categorical_values") |>
  distinct(column, value)
```

`cover_code` has a value of `5`, which is not in the Braun-Blanquet scale
controlled vocabulary (`0`, `0.1`, `0.5`, `1`, `2`, `3`, `4`). `NA` values
are ignored by the categorical check.

### Missing values check

The missing values check evaluates whether columns contain `NA` where the
schema prohibits or discourages it. Each column has one of three rules:

- **`"enforce"`** — `NA` is a **fail**; the column must always be populated.
- **`"warn"`** — `NA` is a **warn**; missing data is unusual but not always
  wrong (e.g., optional or method-dependent fields).
- **`"allow"`** (or no rule) — `NA` is acceptable; column is not checked.

```{r miss-check}
result |>
  filter(check == "qc_check_missing_values")
```

Row 167 has a missing `percent_cover` value. Because `percent_cover` is a
`"warn"` column (cover is expected but `NA` is acceptable when a different
cover method is used), this produces a `"warn"` row rather than a `"fail"`.
The `"enforce"` columns — including `site_name`, `scientific_name`, and
`sample_collection_date` — are fully populated, so they produce no rows.

### Numeric ranges check

The numeric ranges check validates that numeric columns fall within bounds
defined in `marinegeo_metadata$numeric_ranges`. Bounds can be set independently
for **fail** (hard limit) and **warn** (soft limit), and each column uses either
`"inclusive"` (`<`, `>`) or `"exclusive"` (`<=`, `>=`) comparisons.

```{r range-check}
result |>
  filter(check == "qc_check_numeric_ranges")
```

All values in this dataset are within range, so the check contributes no rows.
`point_count` is all `NA` (the dataset uses Braun-Blanquet codes rather than
point counts), so there are no values to evaluate — `NA` cells are silently
skipped.

### Lookup values check

The lookup values check validates that `partner_code`, `site_name`, `site_code`,
and `scientific_name` values are recognized in the MarineGEO global entity
registries (`marinegeo_metadata$partner_codes`, `$site_codes`,
`$observation_lookup`). `NA` values are always ignored. This check runs
automatically whenever any of these columns are present in `data` — it is not
gated by `table_id`.

```{r lookup-check}
result |>
  filter(check == "qc_check_lookup_values")
```

All `partner_code`, `site_name`, and `scientific_name` values in
`seagrass_cover_example` are recognized in the registries.

---

## Extracting `[row, col]` coordinates

Cell-level issues carry both `row` and `col_index` — the exact `[row, col]`
position of each violating cell in the original data frame. Both are 1-based
(R convention).

```{r row-col-coords}
df_bad <- data.frame(
  site_name = c("Site A", "Site B", "INVALID", "Site A"),
  cover_method = c("point count", "UNKNOWN", "point count", "Brawny-Blanquet"),
  stringsAsFactors = FALSE
)

rules_demo <- data.frame(
  column_name = c("site_name", "site_name", "cover_method", "cover_method"),
  value = c("Site A", "Site B", "point count", "Braun-Blanquet"),
  stringsAsFactors = FALSE
)

result_demo <- qc_check_categorical_values(df_bad, rules_demo)
result_demo
```

`col_index` is the column's position in the original data frame
(`which(colnames(data) == column)`), so it remains correct when `data`
contains columns that are not covered by any rules. To format these as compact
`[row, col]` coordinate strings:

```{r row-col-format}
result_demo |>
  mutate(coordinate = paste0("[", row, ", ", col_index, "]")) |>
  select(coordinate, column, value)
```

---

## Running individual checks

The individual check functions can be called directly with a custom schema,
without needing a registered `table_id`. Each returns a `qc_issues` table.

### `qc_check_columns()`

```{r check-columns-standalone}
qc_check_columns(
  data = seagrass_cover_example,
  expected_columns = c(
    "partner_code",
    "site_name",
    "sample_collection_date",
    "scientific_name",
    "percent_cover"
  )
)
```

### `qc_check_data_types()`

```{r check-types-standalone}
qc_check_data_types(
  data = seagrass_cover_example,
  type_map = c(
    partner_code = "STRING",
    sample_collection_date = "DATE",
    transect = "TINYINT",
    percent_cover = "DOUBLE"
  )
)
```

### `qc_check_missing_values()`

Supply a `rules` data frame with `column_name` and `missing_values` columns.
Only `"enforce"` and `"warn"` rows are evaluated; `"allow"` rows are ignored.

```{r check-miss-standalone}
df_survey <- data.frame(
  site_name = c("Reef A", NA, "Reef C"),
  scientific_name = c("Zostera marina", "Halodule wrightii", "Zostera marina"),
  percent_cover = c(45, 20, NA),
  notes = c(NA, NA, "Dense canopy"),
  stringsAsFactors = FALSE
)

miss_rules <- data.frame(
  column_name = c("site_name", "scientific_name", "percent_cover", "notes"),
  missing_values = c("enforce", "enforce", "warn", "allow"),
  stringsAsFactors = FALSE
)

qc_check_missing_values(df_survey, miss_rules)
```

`site_name` is `"enforce"` and has an `NA` in row 2 (a `"fail"` row);
`percent_cover` is `"warn"` and its `NA` produces a `"warn"` row. The `notes`
column is `"allow"` and is not evaluated. The overall status is `"fail"`
because a fail-level row is present.

### `qc_check_numeric_ranges()`

Supply a `rules` data frame with one row per column and columns `column_name`,
`min_fail`, `max_fail`, `min_warn`, `max_warn`, and `range_type`.

```{r check-range-standalone}
df_obs <- data.frame(
  percent_cover = c(10, 55, 105, -3),
  depth_m = c(1.0, 4.5, 28.0, 3.2)
)

range_rules <- data.frame(
  column_name = c("percent_cover", "depth_m"),
  min_fail = c(0, NA),
  max_fail = c(100, NA),
  min_warn = c(NA, 0),
  max_warn = c(80, 20),
  range_type = c("inclusive", "inclusive"),
  stringsAsFactors = FALSE
)

qc_check_numeric_ranges(df_obs, range_rules)
```

Two `percent_cover` values breach the fail threshold (`105 > 100`, `-3 < 0`).
One value (`55`) exceeds the warn threshold (`> 80`) but is within the fail
bounds, so it is recorded as `"warn"`. `depth_m` has only a warn upper bound of
20 m, and row 3 (28.0 m) triggers it. Fail takes priority: a row that breaches
a fail threshold is not additionally flagged as a warn.

### `qc_check_lookup_values()`

Pass a named list where each name is a column to validate and each value is
the character vector of recognized entries. Columns in `lookups` that are
absent from `data` are silently skipped.

```{r check-lookup-standalone}
df_lookup <- data.frame(
  partner_code = c("USA-IRL", "UNKNOWN-XYZ", "AUS-GBR"),
  scientific_name = c("Zostera marina", "Fake species", "Halodule wrightii"),
  stringsAsFactors = FALSE
)

# In practice these vectors come from marinegeo_metadata; abbreviated here
partner_codes_registry <- c("USA-IRL", "AUS-GBR", "CAN-BCI", "MEX-ALV")
scientific_name_registry <- c(
  "Zostera marina",
  "Halodule wrightii",
  "Thalassia testudinum",
  "Cymodocea nodosa"
)

qc_check_lookup_values(
  data = df_lookup,
  lookups = list(
    partner_code = partner_codes_registry,
    scientific_name = scientific_name_registry
  )
)
```

`"UNKNOWN-XYZ"` is not in the partner codes registry and `"Fake species"` is
not in the species registry, so both rows are flagged; each row's `message`
names the registry consulted. When called via `qc_run()`, the full registries
from `marinegeo_metadata` are used automatically.

---

## Reading from a file

`qc_run()` can read directly from a CSV or XLSX file path, avoiding the need
to pre-load data.

```{r file-path, eval = FALSE}
# CSV
result_csv <- qc_run(
  "path/to/seagrass_cover.csv",
  table_id = "seagrass-cover-monitoring-v1"
)

# Excel (specify sheet if needed)
result_xlsx <- qc_run(
  "path/to/seagrass_cover.xlsx",
  table_id = "seagrass-cover-monitoring-v1",
  sheet = 1L
)
```
