---
title: "Running QC checks on MarineGEO data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running QC checks on MarineGEO data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(marinegeo.utils)
library(dplyr)
```

This vignette demonstrates the `marinegeo.utils` QC functions using
`seagrass_cover_example`, a multi-year seagrass percent cover dataset from
Indian River Lagoon, Florida (`partner_code = "USA-IRL"`).

There are two ways to run QC checks:

1. **`qc_run()`** — single entry point that runs all applicable tests for a
   known `table_id`, dispatching based on metadata.
2. **Individual check functions** — `qc_check_columns()`,
   `qc_check_data_types()`, `qc_check_categorical_values()`,
   `qc_check_missing_values()`, `qc_check_numeric_ranges()`, and
   `qc_check_lookup_values()` — for targeted checks against a custom schema.

---

## Running all checks with `qc_run()`

`qc_run()` accepts a data frame (or file path) and a `table_id`. It queries
the internal `marinegeo_metadata` object to determine which columns are
expected, what data types they should have, and which columns have controlled
vocabularies — then runs all applicable tests.

```{r load-data}
data(seagrass_cover_example)
```

```{r qc-run}
result <- qc_run(
  seagrass_cover_example,
  table_id = "seagrass-cover-monitoring-v1"
)
```

The return value is a named list with four top-level elements:

```{r result-structure}
names(result)
```

`status` is the worst status across all tests (`"fail"` > `"warn"` > `"pass"`).
`n_rows` is the number of rows validated. `tests` is a named list with one
entry per test that was run.

```{r top-level}
cat("Overall status:", result$status, "\n")
cat("Rows validated:", result$n_rows, "\n")
cat("Tests run:", paste(names(result$tests), collapse = ", "), "\n")
```

---

## Summarizing results with `utl_qc_summarize()`

The nested list returned by `qc_run()` is well-structured for programmatic
use but can be hard to scan. `utl_qc_summarize()` flattens it into two tidy
data frames: a one-row-per-test **summary** and a combined **failures** table.

```{r summarize-both}
summary_out <- utl_qc_summarize(result)
names(summary_out)
```

### Summary table

One row per test with columns `test`, `status`, `message`, and `n_failures`.
`n_failures` is `0L` for passing tests, the row count of `$failures` for
non-passing tests, or `NA_integer_` when `detail = FALSE` was used.

```{r summarize-summary}
knitr::kable(summary_out$summary)
```

### Failures table

All per-test failure rows bound into a single data frame with a leading `test`
column identifying the source check. Columns not present in a given test's
failures are filled with `NA`. Returns a zero-row data frame when all tests
pass.

```{r summarize-failures}
knitr::kable(summary_out$failures)
```

Use `type` to request only one of the two outputs:

```{r summarize-type}
# summary only
utl_qc_summarize(result, type = "summary")$summary |>
  knitr::kable()
```

### With `detail = FALSE`

When `qc_run()` is called with `detail = FALSE`, `$failures` is `NULL` for
every test. `utl_qc_summarize()` handles this gracefully — `n_failures` is
`NA_integer_` for any non-passing test, and the failures table is zero-row.

```{r summarize-no-detail}
result_nd <- qc_run(
  seagrass_cover_example,
  table_id = "seagrass-cover-monitoring-v1",
  detail = FALSE
)

utl_qc_summarize(result_nd, type = "summary")$summary |>
  knitr::kable()
```

---

## Inspecting individual test results

Each element of `result$tests` has the same structure: `test`, `status`,
`message`, `summary`, and `failures`.

### Column check

```{r column-check}
col_result <- result$tests$qc_check_columns

cat("Status:", col_result$status, "\n")
cat("Message:", col_result$message, "\n")
```

```{r column-failures}
col_result$failures
```

The example dataset is missing the `taxonomic_id` column. This column is
optional in practice (it is populated after a taxonomy join), but the schema
defines it as expected, so the check flags it as a failure.

### Data type check

```{r type-check}
type_result <- result$tests$qc_check_data_types

cat("Status:", type_result$status, "\n")
cat("Message:", type_result$message, "\n")
knitr::kable(type_result$summary)
```

All 16 present columns have the expected data types.

### Categorical values check

```{r cat-check}
cat_result <- result$tests$qc_check_categorical_values

cat("Status:", cat_result$status, "\n")
cat("Message:", cat_result$message, "\n")
knitr::kable(cat_result$summary)
```

The `summary` shows how many invalid values were found per column. Use
`failures` to see the specific rows:

```{r cat-failures-cover-code}
# Unique invalid values in cover_code
cat_result$failures |>
  filter(column_name == "cover_code") |>
  distinct(value)
```

`cover_code` has a value of `5`, which is not in the Braun-Blanquet scale
controlled vocabulary (`0`, `0.1`, `0.5`, `1`, `2`, `3`, `4`). `NA` values
are ignored by the categorical check.

### Missing values check

The missing values check evaluates whether columns contain `NA` where the
schema prohibits or discourages it. Each column has one of three rules:

- **`"enforce"`** — `NA` is a **fail**; the column must always be populated.
- **`"warn"`** — `NA` is a **warn**; missing data is unusual but not always
  wrong (e.g., optional or method-dependent fields).
- **`"allow"`** (or no rule) — `NA` is acceptable; column is not checked.

```{r miss-check}
miss_result <- result$tests$qc_check_missing_values

cat("Status:", miss_result$status, "\n")
cat("Message:", miss_result$message, "\n")
```

```{r miss-summary}
knitr::kable(miss_result$summary)
```

Row 167 has a missing `percent_cover` value. Because `percent_cover` is a
`"warn"` column (cover is expected but `NA` is acceptable when a different
cover method is used), this produces a `warn` rather than a `fail`. The eight
`"enforce"` columns — including `site_name`, `scientific_name`, and
`sample_collection_date` — are fully populated in this dataset.

```{r miss-failures}
miss_result$failures
```

### Numeric ranges check

The numeric ranges check validates that numeric columns fall within bounds
defined in `marinegeo_metadata$numeric_ranges`. Bounds can be set independently
for **fail** (hard limit) and **warn** (soft limit), and each column uses either
`"inclusive"` (`<`, `>`) or `"exclusive"` (`<=`, `>=`) comparisons.

For `seagrass-cover-monitoring-v1`, three columns have range rules:

| Column | `min_fail` | `max_fail` | `min_warn` | `max_warn` | `range_type` |
|---|---|---|---|---|---|
| `transect` | — | — | 1 | 3 | inclusive |
| `quadrat` | 0 | — | 0 | 50 | inclusive |
| `point_count` | 10 | — | — | 81 | inclusive |

```{r range-check}
range_result <- result$tests$qc_check_numeric_ranges

cat("Status:", range_result$status, "\n")
cat("Message:", range_result$message, "\n")
knitr::kable(range_result$summary)
```

All values in this dataset are within range. `point_count` is all `NA` (the
dataset uses Braun-Blanquet codes rather than point counts), so there are no
values to evaluate — `NA` cells are silently skipped.

### Lookup values check

The lookup values check validates that `partner_code`, `site_name`, `site_code`,
and `scientific_name` values are recognized in the MarineGEO global entity
registries (`marinegeo_metadata$partner_codes`, `$site_codes`,
`$observation_lookup`). `NA` values are always ignored. This check runs
automatically whenever any of these columns are present in `data` — it is not
gated by `table_id`.

```{r lookup-check}
lookup_result <- result$tests$qc_check_lookup_values

cat("Status:", lookup_result$status, "\n")
cat("Message:", lookup_result$message, "\n")
knitr::kable(lookup_result$summary)
```

All `partner_code`, `site_name`, and `scientific_name` values in
`seagrass_cover_example` are recognized in the registries.

---

## Extracting `[row, col]` coordinates from categorical failures

The `failures` data frame returned by `qc_check_categorical_values()` includes
both `row_index` and `col_index` — the exact `[row, col]` position of each
violating cell in the original data frame. Both are 1-based (R convention).

```{r row-col-coords}
df_bad <- data.frame(
  site_name = c("Site A", "Site B", "INVALID", "Site A"),
  cover_method = c("point count", "UNKNOWN", "point count", "Brawny-Blanquet"),
  stringsAsFactors = FALSE
)

rules_demo <- data.frame(
  column_name = c("site_name", "site_name", "cover_method", "cover_method"),
  value = c("Site A", "Site B", "point count", "Braun-Blanquet"),
  stringsAsFactors = FALSE
)

result_demo <- qc_check_categorical_values(df_bad, rules_demo)

knitr::kable(result_demo$failures)
```

`col_index` is the column's position in the original data frame
(`which(colnames(data) == column_name)`), so it remains correct when `data`
contains columns that are not covered by any rules.

To format these as compact `[row, col]` coordinate strings:

```{r row-col-format}
result_demo$failures |>
  mutate(coordinate = paste0("[", row_index, ", ", col_index, "]")) |>
  select(coordinate, column_name, value)
```

## Running individual checks

The individual check functions can be called directly with a custom schema,
without needing a registered `table_id`.

### `qc_check_columns()`

```{r check-columns-standalone}
result_cols <- qc_check_columns(
  data = seagrass_cover_example,
  expected_columns = c(
    "partner_code",
    "site_name",
    "sample_collection_date",
    "scientific_name",
    "percent_cover"
  )
)

cat("Status:", result_cols$status, "\n")
cat("Message:", result_cols$message, "\n")
```

### `qc_check_data_types()`

```{r check-types-standalone}
result_types <- qc_check_data_types(
  data = seagrass_cover_example,
  type_map = c(
    partner_code = "STRING",
    sample_collection_date = "DATE",
    transect = "TINYINT",
    percent_cover = "DOUBLE"
  )
)

cat("Status:", result_types$status, "\n")
cat("Message:", result_types$message, "\n")
```

### `qc_check_categorical_values()`

```{r check-cat-standalone}
rules <- data.frame(
  column_name = c("cover_method", "cover_method", "cover_method"),
  value = c("Braun-Blanquet", "point count", "percent bins")
)

result_cat <- qc_check_categorical_values(
  data = seagrass_cover_example,
  rules = rules
)

cat("Status:", result_cat$status, "\n")
cat("Message:", result_cat$message, "\n")
knitr::kable(result_cat$summary)
```

### `qc_check_missing_values()`

Supply a `rules` data frame with `column_name` and `missing_values` columns.
Only `"enforce"` and `"warn"` rows are evaluated; `"allow"` rows are ignored.

```{r check-miss-standalone}
df_survey <- data.frame(
  site_name = c("Reef A", NA, "Reef C"),
  scientific_name = c("Zostera marina", "Halodule wrightii", "Zostera marina"),
  percent_cover = c(45, 20, NA),
  notes = c(NA, NA, "Dense canopy"),
  stringsAsFactors = FALSE
)

miss_rules <- data.frame(
  column_name = c("site_name", "scientific_name", "percent_cover", "notes"),
  missing_values = c("enforce", "enforce", "warn", "allow"),
  stringsAsFactors = FALSE
)

result_miss <- qc_check_missing_values(df_survey, miss_rules)

cat("Status:", result_miss$status, "\n")
cat("Message:", result_miss$message, "\n")
knitr::kable(result_miss$summary)
```

`site_name` is `"enforce"` and has an `NA` in row 2, so the check fails.
`percent_cover` is `"warn"` and also has an `NA`, but because a fail-level
violation was found first, the overall status is `"fail"`. The `notes` column
is `"allow"` and is not evaluated at all.

```{r check-miss-failures}
knitr::kable(result_miss$failures)
```

### `qc_check_numeric_ranges()`

Supply a `rules` data frame with one row per column and columns `column_name`,
`min_fail`, `max_fail`, `min_warn`, `max_warn`, and `range_type`.

```{r check-range-standalone}
df_obs <- data.frame(
  percent_cover = c(10, 55, 105, -3),
  depth_m = c(1.0, 4.5, 28.0, 3.2)
)

range_rules <- data.frame(
  column_name = c("percent_cover", "depth_m"),
  min_fail = c(0, NA),
  max_fail = c(100, NA),
  min_warn = c(NA, 0),
  max_warn = c(80, 20),
  range_type = c("inclusive", "inclusive"),
  stringsAsFactors = FALSE
)

result_range <- qc_check_numeric_ranges(df_obs, range_rules)

cat("Status:", result_range$status, "\n")
cat("Message:", result_range$message, "\n")
knitr::kable(result_range$summary)
```

Two `percent_cover` values breach the fail threshold (`105 > 100`,
`-3 < 0`). One value (`55`) exceeds the warn threshold (`> 80`) but is within
the fail bounds, so it is recorded as `"warn"`. `depth_m` has no fail
thresholds — only a warn upper bound of 20 m — and row 3 (28.0 m) triggers it.

```{r check-range-failures}
knitr::kable(result_range$failures)
```

`severity` distinguishes fail-level breaches from warn-level ones. Fail takes
priority: a row that breaches a fail threshold is not additionally flagged as
a warn, even if it also exceeds the warn bound.

### `qc_check_lookup_values()`

Pass a named list where each name is a column to validate and each value is
the character vector of recognized entries. Columns in `lookups` that are
absent from `data` are silently skipped.

```{r check-lookup-standalone}
df_lookup <- data.frame(
  partner_code = c("USA-IRL", "UNKNOWN-XYZ", "AUS-GBR"),
  scientific_name = c("Zostera marina", "Fake species", "Halodule wrightii"),
  stringsAsFactors = FALSE
)

# In practice these vectors come from marinegeo_metadata; abbreviated here
partner_codes_registry <- c("USA-IRL", "AUS-GBR", "CAN-BCI", "MEX-ALV")
scientific_name_registry <- c(
  "Zostera marina",
  "Halodule wrightii",
  "Thalassia testudinum",
  "Cymodocea nodosa"
)

lookup_result_standalone <- qc_check_lookup_values(
  data = df_lookup,
  lookups = list(
    partner_code = partner_codes_registry,
    scientific_name = scientific_name_registry
  )
)

cat("Status:", lookup_result_standalone$status, "\n")
cat("Message:", lookup_result_standalone$message, "\n")
knitr::kable(lookup_result_standalone$summary)
```

```{r check-lookup-failures}
knitr::kable(lookup_result_standalone$failures)
```

`"UNKNOWN-XYZ"` is not in the partner codes registry and `"Fake species"` is
not in the species registry, so both rows are flagged. `lookup_source`
identifies which registry was checked for each column. When called via
`qc_run()`, the full registries from `marinegeo_metadata` are used
automatically.

---

## Using `detail = FALSE` in automated pipelines

In automated pipelines where row-level detail is unnecessary (and memory use
matters), pass `detail = FALSE`. All tests still run and statuses are still
reported, but `failures` is `NULL` for every test.

```{r detail-false}
result_pipeline <- qc_run(
  seagrass_cover_example,
  table_id = "seagrass-cover-monitoring-v1",
  detail = FALSE
)

cat("Overall status:", result_pipeline$status, "\n")

# failures are suppressed
lapply(result_pipeline$tests, function(t) {
  cat(t$test, "-", t$status, "| failures:", is.null(t$failures), "\n")
})
```

---

## Reading from a file

`qc_run()` can read directly from a CSV or XLSX file path, avoiding the need
to pre-load data.

```{r file-path, eval = FALSE}
# CSV
result_csv <- qc_run(
  "path/to/seagrass_cover.csv",
  table_id = "seagrass-cover-monitoring-v1"
)

# Excel (specify sheet if needed)
result_xlsx <- qc_run(
  "path/to/seagrass_cover.xlsx",
  table_id = "seagrass-cover-monitoring-v1",
  sheet = 1L
)
```
