---
title: "Working with MarineGEO scientific IDs, taxonomy and functional groups"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with MarineGEO scientific IDs, taxonomy and functional groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(marinegeo.utils)
```

This vignette demonstrates how to interact with taxonomic and functional group functions using `seagrass_cover_example`, a multi-year seagrass percent-cover dataset from Indian River Lagoon, Florida (`partner_code = "USA-IRL"`). Load the dataset, which is included in the R package `marinegeo.utils`:

```{r load-data}
data(seagrass_cover_example)

knitr::kable(head(
  seagrass_cover_example |>
    dplyr::select(
      site_name,
      sample_collection_date,
      transect,
      quadrat,
      scientific_name,
      percent_cover
    )
))

```

The dataset contains observations recorded at either a species- or functional group level, which is indicated by the column `scientific_name`. Each `scientific_name` value is associated with a standardized identifier, `scientific_id`. `scientific_id` is the primary identifier linking observations across MarineGEO taxonomic and functional group metadata tables. It is either an Aphia ID (e.g. `"urn:lsid:marinespecies.org:taxname:123"`) for species/taxonomic rank observations, a functional group ID (e.g. `"FUNCTIONAL:SEAGRASS"`) for observations recorded at a functional group level, or a temporary undefined label (`"UNDEFINED:SOMETHINGNOVEL"``). 

The identifier allows scientific names such as "Gastropoda sp A", "Gastropoda sp B", and "Gastropoda sp C" to all be linked to the same `scientific_id`, "urn:lsid:marinespecies.org:taxname:101", and thus any taxonomic or functional group relationships associated with that ID, while preserving the original abbreviations in the `scientific_name` column.

The functions below retrieves `scientific_id` values.

---

## Review `scientific_id` for each observation

`utl_mg_get_scientific_id()` matches each provided `scientific_name` value against the
`observation_lookup` table bundled with the package and returns a `scientific_id`
value. Input values that are not in the lookup receive `NA` and trigger a
warning.

```{r join-id}

# Run the function with a vector of strings 
utl_mg_get_scientific_id("Thalassia testudinum")

# Here "Gastropoda C" doesn't follow guidance on formatting: an sp or sp. is required between the species identifier and "Gastropoda" for proper matching to a scientific ID. 
utl_mg_get_scientific_id(c("Gastropoda sp A", "Gastropoda sp. B", "Gastropoda", "Gastropoda C"))

# Use the function within a `mutate()` call
# How many distinct names were matched?
knitr::kable(
  seagrass_cover_example |>
    dplyr::mutate(scientific_id = utl_mg_get_scientific_id(scientific_name)) |>
    dplyr::distinct(scientific_name, scientific_id) |>
    dplyr::arrange(scientific_name)
)

```

## Retrieve functional group memberships

MarineGEO maintains nested functional group and taxonomic lookup tables. Taxa and functional/morphological groups are arranged hierarchically, allowing child elements to inherit all of the groups above them. `utl_mg_get_functional_groups()` returns every parent from the broadest group (e.g. "Biota") down for each queried ID. Here we see what groups are associated with Halodule wrightii (APHIA:208925) and Caulerpa sertularioides (APHIA:144474).

```{r functional-groups}
species_ids <- c("urn:lsid:marinespecies.org:taxname:208925", "urn:lsid:marinespecies.org:taxname:144474")

fg <- utl_mg_get_functional_groups(
  species_ids,
  functional_group_tree = "vegetation"
)
knitr::kable(fg)
```

The result is a long-format data frame with one row per
(`scientific_id`, group) pair. The columns are:

| Column | Description |
|---|---|
| `scientific_id` | The queried identifier |
| `group_id` | ID of the group (parent node) |
| `group_name` | Display name of the group (parent node) |

You can also use the function on functional groups, as well:

```{r fg-functional-input}

# Find the parent groups of "Seagrass"
seagrass_id <- "FUNCTIONAL:SEAGRASS"
utl_mg_get_functional_groups(seagrass_id, "vegetation")

```

---

## Assign functional groups to data
While you could filter and join the table above to a data table in order to group by functional groups to do some type of calculation, the `utl_mg_assign_functional_group()` function allows to you to assign functional groups of your choice within a mutate call. For instance, let's create a new column, `functional_group`, and assign observations to either "Seagrass" or "Algae". Then we'll generate a timeseries plot based on those groups and the percent cover variable. 

```{r summary, fig.width = 8}

df_functional_groups <- seagrass_cover_example |>
  dplyr::mutate(
    functional_group = utl_mg_assign_functional_groups(
      fg_tree = "vegetation",
      fg_labels = c("Seagrass", "Algae"),
      scientific_names = scientific_name
    )
  )

knitr::kable(
  df_functional_groups |>
    dplyr::select(
      scientific_name,
      functional_group
    ) |>
    dplyr::distinct()
)

df_functional_groups |>
  dplyr::mutate(year = lubridate::year(sample_collection_date)) |>
  dplyr::filter(!is.na(functional_group)) |>
  dplyr::group_by(site_name, year, functional_group) |>
  dplyr::summarize(percent_cover = mean(percent_cover, na.rm = T)) |>
  ggplot2::ggplot(ggplot2::aes(
    x = year,
    y = percent_cover,
    color = site_name
  )) +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(ggplot2::vars(functional_group))

```
